---
title: "Predict Popularity for Top 50 Songs"
author: "Catalina Zavala"
format: html
self-contained: true
---

## Goal: Create a Model Predicting Song Popularity

Let's read in the data and do some exploratory data analysis:

```{r, echo=FALSE, message=FALSE}
require("knitr")
opts_knit$set(root.dir = "~/R projects")

library(kableExtra)

```

```{r message=FALSE, warning=FALSE}
# R packages that will be helpful for EDA
library(dplyr)
library(psych)
library(DescTools)
# R packages for plotting 
library(ggplot2)
library(ggthemes)
library(gridExtra)

# Read in data file 
pop.df <-read.csv('top50.csv', header = TRUE, sep=',')

knitr::kable(head(pop.df))

# cleaning up variable names
names(pop.df) <- c('X','Track.Name','Artist.Name','Genre','Beats.Per.Minute',
                       'Energy','Danceability','Loudness.dB','Liveness',
                       'Valence','Length','Acousticness','Speechiness',
                       'Popularity')

#removing unlabeled variable 
pop.df <- pop.df %>% select(-X)
```

```{r check data}
# check for missingness
any(is.na(pop.df))

# checking file structure
str(pop.df)
```

```{r descriptives, warning=FALSE}
knitr::kable(psych::describe(pop.df[ ,4:13], check=TRUE), digits=2)
```

We will want to visualize the distribution of all the variables/features, so let's set up a function for quickly plotting histograms. 
```{r histogram function}

# Creating a function to plot histograms 
# including overlay of normal (theoretical) density curve 
hist.plots <- function(df, var, var.name) {
  p1 <- ggplot(df, aes(x = var)) +        
        geom_histogram(aes(y = ..density..), bins = 10) +
        xlab(var.name) + 
        stat_function(fun = dnorm,
            args = list(mean = mean(var),
                               sd = sd(var)),
            col = "#a80226", size = 1)  
  }
```

Let's use Tukey's method for identifying outliers, which will correspond with data visualization in a boxplot e.g. indicating values outside the interquartile range (IQR)
```{r Popularity outliers}

DescTools::Outlier(pop.df$Popularity)

#looking at distribution of Popularity variable 
# and annotate outliers in ggplot boxplot
pop.hist <- hist.plots(pop.df, pop.df$Popularity, "Popularity") +
            scale_x_continuous(limits=c(60, 100))
pop.box <-  ggplot(pop.df, aes(x=Popularity)) + 
            geom_boxplot(outlier.colour = "red") +
            scale_x_continuous(limits=c(60, 100)) +
            annotate(geom="text", x=70, y=.04, label="70", color="red") +
            annotate(geom="text", x=78, y=.04, label="78", color="red")

gridExtra::grid.arrange(pop.hist, pop.box, nrow=2)

#based on interquartile range, capping/flooring outliers in Popularity 
IQR.low = stats::quantile(pop.df$Popularity, .25) - 1.5*stats::IQR(pop.df$Popularity)
pop.cap <- round(IQR.low)
pop.df$Popularity <- DescTools::Winsorize(pop.df$Popularity, minval = 79, maxval = 95)

psych::describe(pop.df$Popularity)

```

To reduce left skewness, power transformations, such as squares or cubes are recommended.
reference: <http://fmwww.bc.edu/repec/bocode/t/transint.html>
```{r clean and transform Popularity, fig.asp = 0.8, fig.width = 8}

#transform
pop.df$Popularity = (pop.df$Popularity)^3
pop.df$Popularity = (pop.df$Popularity)/1e4 #put back onto a similar scale 
psych::describe(pop.df$Popularity)
```

Let's take a look at the correlations of Population and the available features 
```{r correlations, message=FALSE, warning=FALSE, fig.asp = 0.8, fig.width = 8}

library(corrplot)

#selecting numeric features/variables 
cor.data <- stats::cor(pop.df[, sapply(pop.df, is.numeric)])
corrplot(cor.data,method='number')

```

## Distributions (histograms) of potential features 
```{r feature histograms, fig.asp = 0.8, fig.width = 8}
p1 <- hist.plots(pop.df, pop.df$Acousticness, "Acousticness")
p2 <- hist.plots(pop.df, pop.df$Beats.Per.Minute, "Beats.Per.Minute")
p3 <- hist.plots(pop.df, pop.df$Danceability, "Danceability")
p4 <- hist.plots(pop.df, pop.df$Energy, "Energy")
p5 <- hist.plots(pop.df, pop.df$Length, "Length")
p6 <- hist.plots(pop.df, pop.df$Liveness, "Liveness")
p7 <- hist.plots(pop.df, pop.df$Loudness.dB, "Loudness.dB")
p8 <- hist.plots(pop.df, pop.df$Speechiness, "Speechiness")
p9 <- hist.plots(pop.df, pop.df$Valence, "Valence")

gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, nrow = 3)
```
Many of the features have fairly skewed distributions, so let's automate the process for finding the best transformation for each variable.  
Method:
*Raymaekers, J., & Rousseeuw, P. J. (2021). Transforming variables to central normality. Machine Learning, 1-23.* available at: <https://link.springer.com/article/10.1007/s10994-021-05960-5>
```{r transform features, message=FALSE}

library(cellWise)

#subsetting features to be transformed
trans.am<- subset(pop.df, select= c('Acousticness','Beats.Per.Minute','Danceability', 
                                    'Energy','Length','Liveness','Loudness.dB', 
                                    'Valence'))

# type="bestObj" will choose between 
# Box-Cox or Yeo-Johnson transformation for each variable
# robust=TRUE, Reweighted Maximum Likelihood method is used
# to choose λ for each variable transformation 
auto.pontiac <- cellWise::transfo(trans.am, type = "bestObj", robust=TRUE)

# Note: First run of the transfo function returned the following warning: 
# It is recommended not to transform the variable Speechiness or to do so manually.

#export transformed scores
firebird.df <-as.data.frame(auto.pontiac$Xt)

#export info on what type of transformation was used for each variable
fire.bird <-as.data.frame(auto.pontiac$ttypes)
names(fire.bird)[names(fire.bird)=="auto.pontiac$ttypes"] <- "ttype"

# collect info on how each variable was transformed 
fire.bird$var.name <- c('Acousticness','Beats.Per.Minute','Danceability', 
                        'Energy','Length','Liveness','Loudness.dB', 
                        'Valence')

smash <- function(var,t.type) {
  paste(var," (",t.type,")", sep="")  }

fire.bird$var.t<-smash(fire.bird$var.name,fire.bird$ttype)

#  To reduce right skewness, roots, logarithms, or reciprocals are recommended. 
pop.df$logSpeechiness = log(pop.df$Speechiness)
```

Now we can take a look at the results of the transformations, including whether Box-Cox or Yeo-Johnson transformations were applied. 
```{r look at transformed distributions, fig.asp = 0.8, fig.width = 8}

# Draw histograms for transformed variables 
# Pull feature name and transformation type from fire.bird
tr1 <- hist.plots(firebird.df, firebird.df$Acousticness, fire.bird[1,3])
tr2 <- hist.plots(firebird.df, firebird.df$Beats.Per.Minute, fire.bird[2,3])
tr3 <- hist.plots(firebird.df, firebird.df$Danceability, fire.bird[3,3])
tr4 <- hist.plots(firebird.df, firebird.df$Energy, fire.bird[4,3])
tr5 <- hist.plots(firebird.df, firebird.df$Length, fire.bird[5,3])
tr6 <- hist.plots(firebird.df, firebird.df$Liveness, fire.bird[6,3])
tr7 <- hist.plots(firebird.df, firebird.df$Loudness.dB, fire.bird[7,3])
tr8 <- hist.plots(pop.df, pop.df$logSpeechiness, "Speechiness (Log)")
tr9 <- hist.plots(firebird.df, firebird.df$Valence, fire.bird[8,3])

gridExtra::grid.arrange(tr1, tr2, tr3, tr4, tr5, tr6, tr7, tr8, tr9, nrow = 3)

```

```{r transformed descriptives}
#collect transformed variables into the same data frame
pop.part1 <- subset(pop.df, select = c('logSpeechiness','Popularity'))
pop.dft.num <- cbind(firebird.df,pop.part1)
knitr::kable(describe(pop.dft.num), digits=2)
```

```{r remaining outliers}
#check for outliers
DescTools::Outlier(pop.dft.num$Length)

IQR.low = stats::quantile(pop.dft.num$Length, .25) - 1.5*stats::IQR(pop.dft.num$Length)

# cap any remaining outliers to minimum value to retain all data points
pop.dft.num$Length <- DescTools::Winsorize(pop.dft.num$Length, 
                                           minval =-0.315553293,
                                           maxval = 0.323574509)
```

There is an additional, categorical feature to consider: music genre.
```{r music genre, fig.asp = 0.8, fig.width = 8}

#add categorical variable to data frame
pop.dft <- pop.dft.num
pop.dft$genre <-pop.df[, c(3)]

bp1 <-ggplot2::ggplot(pop.dft, aes(x=genre, y=Popularity, color=genre))   
bp1 <- bp1 + geom_boxplot() + geom_jitter(width=0.10) + theme(legend.position = "none")
bp1 <- bp1 + theme(axis.text.x = element_text(angle = 45, hjust = 1))
bp1

unique(pop.dft$genre)

#re-coding genre to pop vs other
pop.dft$genre.2 <- ifelse(pop.dft$genre == "pop", "pop", 
                    ifelse(pop.dft$genre == "canadian pop", "pop",
                    ifelse(pop.dft$genre == "dance pop", "pop", 
                    ifelse(pop.dft$genre == "electropop", "pop", 
                    ifelse(pop.dft$genre == "pop house", "pop",
                    ifelse(pop.dft$genre == "panamanian pop", "pop", 
                    ifelse(pop.dft$genre == "australian pop", "pop", "other")))))))
```

```{r t test, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
# add t-test results to box plot to include more info on group differences 
stats::t.test(pop.dft$Popularity ~ pop.dft$genre.2)
```

```{r genre boxplot, message=FALSE, warning=FALSE, echo=TRUE, fig.asp = 0.8, fig.width = 8}

bp2 <-ggplot2::ggplot(pop.dft, aes(x=genre.2, y=Popularity, color=genre.2))   
bp2 <- bp2 + geom_boxplot() + geom_jitter(width=0.10) + theme(legend.position = "none")
bp2 <- bp2 + annotate(geom="text", x=1.5, y=89, label="t(40.2) = 2.65, p = 0.01", color="black") 
bp2

```

```{r some maintenance }

#designate new genre variable as factor 
pop.dft$genre.2 <- as.factor(pop.dft$genre.2)
#remove previous version of genre from data frame
pop.dft <-pop.dft[-c(11)]
# double check new file structure 
str(pop.dft)

# get rid of some unnecessary objects
rm(pop.df)
rm(pop.dft.num)
rm(firebird.df)
rm(fire.bird)
rm(pop.part1)
rm(tr1, tr2, tr3, tr4, tr5, tr6, tr7, tr8, tr9)
rm(p1, p2, p3, p4, p5, p6, p7, p8, p9)
```

Now we need to split the data and normalize the variables before training. 
```{r setting up training, message=FALSE, warning=FALSE}
library(caTools)
library(caret) # for z-scoring test & train sets 

# Setting a random seed so results are reproducible
set.seed(123) 

# creating train and test data sets
# Split up the sample, basically randomly assigns a Boolean value to a new column "sample"
sample <- sample.split(pop.dft$Length, SplitRatio = .7 )
# Training Data
train = subset(pop.dft, sample == TRUE)
# Testing Data
test = subset(pop.dft, sample == FALSE)

# save out mean and sd from training data to use for z-scoring
# both train and test data sets 
zscore.train <- preProcess(train, method = c("center", "scale"))
train.z <-predict(zscore.train, train)
test.z <-predict(zscore.train, test)

knitr::kable(psych::describe(train.z, check=TRUE), digits=3)
```

## Lasso Regression
Since sample size is small with quite a few features, let's use Lasso regression to help with variable selection.

```{r lasso, message=FALSE, warning=FALSE, fig.asp = 0.8, fig.width = 8}
#lasso analysis 
library(glmnet)
library(Matrix)

# Prepping data for model. 
# Decided to exclude Beats.Per.Minute as it is fairly correlated with Speechiness (0.56)
x <- data.matrix(train.z[, -c(2,7,10)]) 
y <- train.z$Popularity

# using folds=35 to set up LOOVC 
cv_model <- glmnet::cv.glmnet(x=x, y=y, alpha = 1, folds=35, standardize=FALSE)
print(cv_model)
plot(cv_model)
```

```{r best model}
best_lambda <- cv_model$lambda.min
best_lambda
log(best_lambda)

#save out best model for prediction
best_model <- glmnet::glmnet(x, y, alpha = 1, lambda = best_lambda)
coef(best_model)
```

## Visualization of how each feature performed in the Lasso analysis 
```{r lasso features plot, message=FALSE, fig.asp = 0.8, fig.width = 8}

library(tidyr)
library(ggrepel)

#get data out to plot how features performed in CV
betas = as.matrix(cv_model$glmnet.fit$beta)
lambdas = log(cv_model$lambda)
names(lambdas) = colnames(betas)

as.data.frame(betas) %>% 
  tibble::rownames_to_column("variable") %>% 
  pivot_longer(-variable) %>% 
  mutate(lambda=lambdas[name]) %>% 
  ggplot(aes(x=lambda,y=value,col=variable)) + theme(legend.position = "none") + 
  geom_line() + scale_y_continuous(breaks = seq(-.75, .50, by = .25)) +
  scale_x_continuous(breaks = seq(-1, -8, by = -1)) +
  geom_label_repel(data=~subset(.x,lambda==min(lambda)), 
                   aes(label=variable),nudge_x=-0.5) + 
  annotate(geom="text", x=-7.5, y=.35 , label="df = ", color="red") +
  annotate(geom="text", x=-7, y=.35 , label="9", color="red") + 
  annotate(geom="text", x=-3, y=.35 , label="6", color="red") + 
  annotate(geom="text", x=-2, y=.35 , label="5", color="red") + 
  annotate(geom="text", x=-1, y=.35 , label="0", color="red") +
  geom_vline(xintercept=c(-2.68), linetype="dotted")
```

Let's take a look at R squared in the training set
```{r message=FALSE, warning=FALSE}
#get r squared values for training data
y.pred.train <-stats::predict(best_model, x)

train.sst <- sum((y - mean(y))^2)
train.sse <- sum((y - y.pred.train)^2)

#use SST and SSE to calculate R-Squared
train.rsq <- 1 - train.sse/train.sst
train.Rsq <-round(train.rsq, digits=3)
train.Rsq

```

## Test set performance
```{r test set, message=FALSE, warning=FALSE}
#prep test data set to predict with best model
test.x <- data.matrix(test.z[,-c(2,7,10)])
test.y  <- test.z$Popularity

#use fitted best model to make predictions on test set
y.pred <- stats::predict(best_model, newx = test.x)

#find Test SST and SSE
test.sst <- sum((test.y - mean(test.y))^2)
test.sse <- sum((test.y - y.pred)^2)

#use SST and SSE to calculate Test R-Squared
test.rsq <- 1 - test.sse/test.sst
test.Rsq <-round(test.rsq, digits=3)

#Test MSE
test.int.model.MSE <-round(mean((mean(y) - test.y)^2), digits=3)
test.best.model.MSE <-round(mean((y.pred - test.y)^2), digits=3)

test.samp <-data.frame(test.int.model.MSE, test.best.model.MSE, test.Rsq)

kbl(test.samp) %>%
  kable_paper("hover", full_width = F)

```

